Why robot programming is hard

### 2013-11-30 paper.tex
First pass at rewrite post-talk

\documentclass[10pt,twocolumn]{article}
\usepackage{graphicx}

@@

\title {Why robot programming is hard}
\author {Tessa Lau \\
Affiliation TBD \\
Buenos Aires, Argentina \\
{\tt tlau@tlau.org}
}
\maketitle

@@

Yet the holy grail of personal service robots remains elusive. I believe there are three main reasons. People have a wide variety of tasks for service robots to perform, ranging from folding laundry, to finding lost car keys, to picking up the kids' toys, or doing the dishes and putting them away. They also want robots to perform these tasks in a variety of environments, including cluttered homes and kitchens with different affordances such as modern drawer pulls or faucet handles. Most importantly, these service robots must be easy to use by non-technical people.

We have started to see robots enter the marketplace that address two out of those three hurdles. For example, iRobot's Roomba only does vacuuming, but it does it in a variety of environments and is dead simple to use even by non-technical folks. On the other hand, Rethink Robotics' Baxter robot can be easily programmed to perform a variety of tasks in light industrial settings, but its lack of mobility make it less suitable for the home. As another example, Willow Garage's PR-2 robot has been programmed to perform a variety of household tasks, in a variety of environments, but can only be operated by highly skilled roboticists.

@@

With this goal in mind, personal service robots are more than a consumer toy, but a technology that could changes people's lives for the better. Why, then, are robots still so difficult to use? What makes robots so difficult to control?

@@

\item {\bf Amount of autonomy}: robots are gaining the ability to perform some functions autonomously, such as spatial navigation and object manipulation.  The less autonomy exhibited by the robot, the easier it is to program the robot, because increased autonomy requires the programmer to develop more complex mental models of the robot's perceptions and behavior. Teleoperation interfaces are fairly easy to program because there is generally a one-to-one mapping between the user's inputs and the robot's behavior. Once a robot starts to act autonomously, understanding its behavior (and debugging it!) becomes more challenging.
\end{itemize}

@@

This interface, while well suited for roboticists who need to examine the low-level sensor data and messages being processed by the robot, is not very user-friendly. It also illustrates some of the difficulties inherent in robot programming.

\section{A 3D world}

@@

\section{7-DOF arms}

A different aspect of high-dimensionality comes from the design of robot arms. A solid object in 3D space has six degrees of freedom (DOF): three dimensions to specify its position, and three more to specify its rotation. If a robot arm had only 6-DOF and was trying to execute a specific grasp of a particular object, such as picking up soda can from the side, there would be only a single arm configuration that would allow the gripper (a robotic hand) to pick up the can. In contrast, human arms and advanced robot arms have 7-DOF, which allows them one extra dimension of flexibility in achieving those grasps. To visualize this, imagine grasping a soda can on the table in front of you and then flapping your elbow up and down while keeping the can stationary. That extra degree of freedom enables your arm to take on all those different shapes while still achieving the desired grasp.

Interfaces for robot arm teleoperation typically involve setting the position or angle of each joint in the arm individually. For example, a tablet-style interface might have fourteen buttons, one pair used to increase or decrease the angle of each joint in the arm. That additional degree of freedom makes it easier to identify a valid arm configuration for executing a grasp, since only one of many valid configurations must be found, not the unique one.  However controlling each individual joint position by clicking on buttons is tedious. Additionally, since the allowable joint positions are not completely independent, finding a correct solution is often done through trial and error.

Toolkits for doing automatic motion planning are maturing, which will simplify the problem of robot arm control from how the arm should move in fine-grained detail to specifying only the high-level target position and orientation. Yet the challenge of specifying that 6-DOF target position and orientation still remains, and this will continue to be a challenge for easy-to-use robot interfaces that work with existing 2D displays and input devices.

@@

\section{Uncertainty}

What does this mean for robot programmers? Robot programs must have mechanisms in place to deal with uncertain truth and unpredictable hardware. Robot control loops usually sense the physical hardware during operation to determine whether a motion was carried out successfully. Unlike with demonstration and teleop, where the operator can see if the robot carried out the desired motion and take appropriate action, a robot programmer must think through possible outcomes while writing a program and craft appropriate responses for each eventuality. For example, while instructing a robot to pick up an object, a programmer must specify what to do if the object is not {\em where} it thinks it is, is not {\em what} it thinks it is, or if the robot fails to perform the desired grasp. Should it simply try again? Move a little bit to get a better view and then try again? Ask a human?

In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models and learn to craft fault-tolerant code with frequent error checking, which seems like an heavy burden for casual programmers. Most end user programming systems to date simply ignore the problem of uncertainty. For robot programming to be successful, we will need to develop easily-understandable paradigms for handling uncertainty.

\section{Concurrency}

Robots are complex beings full of many moving parts. For example, Willow Garage's PR-2 has two independently controlled 7-DOF arms, a head that can pan and tilt, a torso that can be raised and lowered, and a base that can move and turn in all directions. Robots are also equipped with a slew of sensors including cameras, depth sensors, laser rangefinders, touch sensors, and feedback about its electrical and motor systems. Sensors and actuators must be operated together in coordination to produce a desired robotic behavior.

Concurrent programming is notoriously difficult even for professional programmers due to the possibility for race conditions and the complexity of reasoning about what the program does. Most robot programs can be simplified by only controlling one moving part at a time: first move the base, then raise the torso, then position the left arm, then position the right arm, finally open the cabinet. For many tasks this is a reasonable solution. However there is a subset of human tasks for which we need multiple parts moving simultaneously, such as picking up a heavy or bulky object with two hands, or turning one's head to look where one is going.

End user programming systems have tried to simplify the problem of concurrency. Some systems visualize the different components on a timeline. This approach is popular with media creation tools such as Apple's GarageBand or Autodesk's 3D Studio Max. A second paradigm, taken by visual programming languages such as Scratch~\cite{scratch}, makes the notion of time a global constant and runs all blocks of user code simultaneously in the virtual environment. More research is needed to discover what paradigm works best for controlling robot parts concurrently.

Even if individual robot parts can be programmed to act in parallel, there will still be a need for reasoning about concurrent robot behavior at a higher level. One form this will take is for background tasks, such as specifying that a robot should try to tidy up the room in between (or during) its other assigned chores. Supporting high level concurrency will introduce new problems, such as task prioritization (is the current task of bringing you a beer more important than cleaning up the wine spill it just detected?).

@@

Robot programming environments are beginning to address all of these challenges and produce useful systems that can be controlled by untrained users. The key tradeoff is simplicity versus expressiveness: in order to make systems that are more tractable for novices, they need to simplify key aspects of the robot's behavior, thereby sacrificing its ability to perform a full range of behavior.

The current generation of commercial home robots illustrates one extreme end of the simplicity versus expressiveness tradeoff. iRobot's successful Roomba line of vacuum-cleaning robots\footnote{http://www.irobot.com/us/learn/home/roomba.aspx} sports a single-button interface that simply says ``Clean''. While the robot will eventually clean your floors, you lose control over how exactly it does so, as it performs what seems like a random walk through your home.

@@

Roboticists are hard at work developing increasingly autonomous robot capabilities for well-specified tasks such as 2D navigation (getting from point A to point B within a building), arm motion planning (moving the gripper to a specified position and orientation while avoiding obstacles)\footnote{MoveIt!, http://moveit.ros.org}, and recognizing and picking up discrete objects on a flat surface. Autonomy is key for enabling robots to automatically perform longer and more complex tasks on their own.

Our approach to designing end-user-programmable interfaces is to craft these autonomous components into a good set of {\em primitives} in an end user programming language, such that a non-expert would not have to ``look inside the box'' to understand how each one works. Roboticists can then implement these primitives using advanced techniques to solve the difficult problems encountered during robot programming. For example, a good autonomous navigation component could incorporate a set of recovery behaviors to try on navigation failure, such as turning around, waiting for people/obstacles to clear out of its path, and backing up and retrying. It could also bake in some concurrent strategies, such as turning the head to look in the direction the robot is going.

@@

We have made use of these autonomous functions to develop a system at Willow Garage called Continuous Ops (short for ``Continuous Operations'') that enables casual users to control our PR-2 robots through a web interface (Figure~\ref{contops}). This web interface, accessible from standard laptops or tablets, provides the ability to tell a PR-2 robot to unplug itself from the power supply, navigate autonomously to anywhere in the buliding, position itself in front of a table, automatically identify graspable objects on that table, and pick up an object.

This interface lies somewhere between teleoperation and programming. It does not require an operator to exert direct control over the robot, but the commands that are given are executed immediately by the robot rather than scheduled to occur later in time. We are also exploring interfaces for sequencing these primitives into scheduled programs, in order to validate that the set of primitives we are developing is complete enough to perform useful tasks.

Another key component of our approach, which we call {\em shared autonomy}, is to fall back on teleoperator assistance when autonomy fails, as it will invariably do for the near future. In our system designs, the teleoperator does not necessarily have to be the person who wrote the program. Instead, we are modeling our system after telephone call centers where a team of trained professionals handle incidents as they arise and get the robots back on track to fulfill their programmed goals. As a stepping-stone towards full autonomy, this approach enables us to start testing end user programmable systems without needing a fully reliable set of autonomous primitives in place to support them.

Over time, as we refine existing autonomous primitives and build new domain-specific primitives, my hope is that we will expand the repertoire of robust end-user-programmable robot behavior. And perhaps that goal of having our own Rosie the Robot in our homes will soon become a reality.

% Acknowledgements
% Andreas Paepke
% Kaijen Hsiao
% Matei Ciocarlie
% Jon Binney
% Maya Cakmak
% anonymous reviewers

>>>>

\documentclass[10pt,twocolumn]{article}
\usepackage{graphicx}
\usepackage{changebar}

@@

\title {Why robot programming is hard}
\author {Tessa Lau \\
Savioke Inc.\\
Buenos Aires, Argentina \\
{\tt tlau@tlau.org}
}
\maketitle

@@

\begin{changebar}
Yet the holy grail of personal service robots remains elusive. I believe there are three main reasons. People have a wide variety of tasks for service robots to perform, ranging from folding laundry, to finding lost car keys, to picking up the kids' toys, or doing the dishes and putting them away. They also want robots to perform these tasks in a variety of environments, including cluttered homes and kitchens with different affordances such as modern drawer pulls or faucet handles. Most importantly, these service robots must be easy to use by non-technical people.

We have started to see robots enter the marketplace that address two out of those three hurdles. For example, iRobot's Roomba only does vacuuming, but it does it in a variety of environments and is dead simple to use even by non-technical folks. On the other hand, Rethink Robotics' Baxter robot can be easily programmed to perform a variety of tasks in light industrial settings, but its lack of mobility make it less suitable for the home. As another example, Willow Garage's PR-2 robot has been programmed to perform a variety of household tasks, in a variety of environments, but can only be operated by highly skilled roboticists.
\end{changebar}

@@

With this goal in mind, personal service robots are more than a consumer toy, but a technology that could changes people's lives for the better.
\begin{changebar}
Why, then, are robots still so difficult to use? What makes robots so difficult to control?

@@

\item {\bf Amount of autonomy}: robots are gaining the ability to perform some functions autonomously, such as spatial navigation and object manipulation.  The less autonomy exhibited by the robot, the easier it is to program the robot, because increased autonomy requires the programmer to develop more complex mental models of the robot's perceptions and behavior. Teleoperation interfaces are fairly easy to program because there is generally a one-to-one mapping between the user's inputs and the robot's behavior. Once a robot starts to act autonomously, understanding its behavior (and debugging it!) becomes more challenging.
\end{itemize}
\end{changebar}

@@

\begin{changebar}
\section{Why robot programming is hard}
\end{changebar}

The Rviz interface, while well suited for roboticists who need to examine the low-level sensor data and messages being processed by the robot, is not very user-friendly. It also illustrates some of the difficulties inherent in robot programming.

\subsection{Concurrency}

\begin{changebar}
Many activities of daily living require coordinated action amongst multiple limbs. Anyone who has lived with a broken arm, for example, understands the difficulty of trying to unscrew jar lids, pick up bulky objects, or chop vegetables with only a single arm. Controlling both arms simultaneously to perform coordinated actions is a matter of concurrent programming.

The problem becomes even more complex with robots that have more moving parts.
\end{changebar}
For example, Willow Garage's PR-2 has two independently controlled 7-DOF arms, a head that can pan and tilt, a torso that can be raised and lowered, and a base that can move and turn in all directions. Robots are also equipped with a slew of sensors including cameras, depth sensors, laser rangefinders, touch sensors, and feedback about its electrical and motor systems. These sensors and actuators must be operated together in coordination to produce a desired robotic behavior.

\begin{changebar}
Concurrent programming is notoriously difficult even for professional programmers due to the possibility for race conditions and the complexity of reasoning about what the program does. Professional programmers learn to use advanced techniques such as multi-threading, event-driven callbacks, and coroutines in order to specify concurrent behavior in programs. Yet these concepts are notoriously difficult to use and debug, even for trained programmers; how can we expect non-technical users to be able to do the same?

In addition to the low level concurrency needed to control multiple robot parts, there is also a need to specify concurrent robot behavior at a higher level. For example, a user may want to give the robot background tasks (such as cleaning or plugging oneself in when the battery is low) which should be scheduled in concert with foreground tasks (such as fetching a drink).
\end{changebar}

%Most robot programs can be simplified by only controlling one moving part at a time: first move the base, then raise the torso, then position the left arm, then position the right arm, finally open the cabinet. For many tasks this is a reasonable solution. However there is a subset of human tasks for which we need multiple parts moving simultaneously, such as picking up a heavy or bulky object with two hands, or turning one's head to look where one is going.

%End user programming systems have tried to simplify the problem of concurrency. Some systems visualize the different components on a timeline. This approach is popular with media creation tools such as Apple's GarageBand or Autodesk's 3D Studio Max. A second paradigm, taken by visual programming languages such as Scratch~\cite{scratch}, makes the notion of time a global constant and runs all blocks of user code simultaneously in the virtual environment. More research is needed to discover what paradigm works best for controlling robot parts concurrently.

\subsection{Uncertainty}

Thus robot programs must have mechanisms in place to deal with uncertain truth and unpredictable hardware. Robot control loops usually sense the physical hardware during operation to determine whether a motion was carried out successfully. If the operator is not physically co-located with the robot, she cannot see whether the robot carried out the desired motion, so she must think through possible outcomes while writing a program and craft appropriate responses for each eventuality. For example, while instructing a robot to pick up an object, a programmer must specify what to do if the object is not {\em where} it thinks it is, is not {\em what} it thinks it is, or if the robot fails to perform the desired grasp. Should it simply try again? Move a little bit to get a better view and then try again? Ask a human?

\begin{changebar}
In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models and craft fault-tolerant code with frequent error checking, which seem like an heavy burden for casual programmers who just want a robot to do something useful in the home.
\end{changebar}

%Most end user programming systems to date simply ignore the problem of uncertainty. For robot programming to be successful, we will need to develop easily-understandable paradigms for handling uncertainty.

\begin{changebar}
\subsection{Commonsense reasoning}

Reasoning about objects in the world requires some amount of common sense which has yet to be programmed consistently into robots. For example, a robot who has been asked to pick up and deliver a coffee cup may well select a grasp that involves dipping its fingers into the liquid and picking up the cup by its side wall. Although the grasp would be valid, this behavior would be judged somewhat undesireable by human observers.

As another example, robots that are capable of automated navigation may fail to observe human norms when navigating through human spaces. The custom of staying to the right when passing in a hallway, for example, can cause confusion if a robot charges down the middle of the hall and does not yield to oncoming passerby.

Endowing computers with common sense is still an active research area. Techniques for making use of common sense reasoning have not yet made it into the vocabulary of professional programmers. Yet these concepts will be taken for granted by end users of robots, who will expect their robots to obey common social norms as they perform their tasks.
\end{changebar}

\subsection{A 3D world}

@@

%\section{7-DOF arms}
%
%A different aspect of high-dimensionality comes from the design of robot arms. A solid object in 3D space has six degrees of freedom (DOF): three dimensions to specify its position, and three more to specify its rotation. If a robot arm had only 6-DOF and was trying to execute a specific grasp of a particular object, such as picking up soda can from the side, there would be only a single arm configuration that would allow the gripper (a robotic hand) to pick up the can. In contrast, human arms and advanced robot arms have 7-DOF, which allows them one extra dimension of flexibility in achieving those grasps. To visualize this, imagine grasping a soda can on the table in front of you and then flapping your elbow up and down while keeping the can stationary. That extra degree of freedom enables your arm to take on all those different shapes while still achieving the desired grasp.
%
%Interfaces for robot arm teleoperation typically involve setting the position or angle of each joint in the arm individually. For example, a tablet-style interface might have fourteen buttons, one pair used to increase or decrease the angle of each joint in the arm. That additional degree of freedom makes it easier to identify a valid arm configuration for executing a grasp, since only one of many valid configurations must be found, not the unique one.  However controlling each individual joint position by clicking on buttons is tedious. Additionally, since the allowable joint positions are not completely independent, finding a correct solution is often done through trial and error.
%
%Toolkits for doing automatic motion planning are maturing, which will simplify the problem of robot arm control from how the arm should move in fine-grained detail to specifying only the high-level target position and orientation. Yet the challenge of specifying that 6-DOF target position and orientation still remains, and this will continue to be a challenge for easy-to-use robot interfaces that work with existing 2D displays and input devices.
%

@@

@@

\begin{changebar}
The field of end user programming~\cite{lieberman-yourwish,dontcheva-nocode} has developed a variety of solutions to enable users of software systems to create or modify software artifacts without having to learn the intricacies of programming.

The key insight for these EUP systems is to reduce the complexity of a programming task through simplification. The three primary methods include (a) visual programming: simplifying or eliminating tricky syntax through graphical representations; (b) domain-specific languages: limiting concepts to only those necessary in a particular domain of use, such as spreadsheet formulas; and (c) programming by demonstration or example: letting users focus on concrete examples rather than difficult abstractions.

However, this simplicity typically comes at the price of expressiveness. By reducing the complexity exposed in an interface, we also reduce the number or variety of functions that can be performed in that interface. In order to make robots that are more tractable for non-technical users, we will have to simplify key aspects of robot behavior, thus sacrificing robots' ability to perform a full range of behavior.
\end{changebar}

%The current generation of commercial home robots illustrates one extreme end of the simplicity versus expressiveness tradeoff. iRobot's successful Roomba line of vacuum-cleaning robots\footnote{http://www.irobot.com/us/learn/home/roomba.aspx} sports a single-button interface that simply says ``Clean''. While the robot will eventually clean your floors, you lose control over how exactly it does so, as it performs what seems like a random walk through your home.

@@

%Roboticists are hard at work developing increasingly autonomous robot capabilities for well-specified tasks such as 2D navigation (getting from point A to point B within a building), arm motion planning (moving the gripper to a specified position and orientation while avoiding obstacles)\footnote{MoveIt!, http://moveit.ros.org}, and recognizing and picking up discrete objects on a flat surface. Autonomy is key for enabling robots to automatically perform longer and more complex tasks on their own.

\begin{changebar}
\subsection{Higher-level primitives}

One way to trade off expressiveness for simplicity is to abstract away details of robot behavior into higher-level primitives. This abstraction can be difficult for roboticists to accept, who have become accustomed to having fine-grained control over robot behavior and error handling. Yet abstracting away these details is going to be critical for end users to be able to develop a working understanding of the robot's functionality.

%Our approach to designing end-user-programmable interfaces is to craft these autonomous components into a good set of {\em primitives} in an end user programming language, such that a non-expert would not have to ``look inside the box'' to understand how each one works.

Roboticists can then implement these primitives using advanced techniques to solve the difficult problems encountered during robot programming.
\end{changebar}
For example, a good autonomous navigation component could incorporate a set of recovery behaviors to try on navigation failure, such as turning around, waiting for people/obstacles to clear out of its path, and backing up and retrying. It could also bake in some concurrent strategies, such as turning the head to look in the direction the robot is going.

@@

We made use of these autonomous functions to develop a system at Willow Garage called Continuous Ops (short for ``Continuous Operations'') that enabled casual users to control PR-2 robots through a web interface (Figure~\ref{contops}).
\begin{changebar}
The interface was inspired by the head of Willow Garage, who noted that even though he had a lab full of robots, he himself wasn't able to show off any of the functionality of his robots to visitors without a roboticist present to operate them.
\end{changebar}

In response, we developed a web interface, accessible from standard laptops or tablets, that provided the ability to tell a PR-2 robot to perform a variety of tasks.  It could unplug itself from the power supply, navigate autonomously to anywhere in the buliding, position itself in front of a table, automatically identify graspable objects on that table, and pick up an object.
\begin{changebar}
The interface enabled non-technical users to activate several core autonomous functions of the PR-2 through an easy-to-use interface.

The drawback of this interface was that the user lost fine-grained control over exactly how the robot performed these tasks. For example, the interface only supported using the robot's right arm to pick up an object; the left arm was unused. It did not allow the user to specify exactly how to pick up an object (e.g. from above or from the side), even though these parameters can be customized at a lower level of control. The challenge in designing a simpler interface is to carefully craft exactly which functionality to expose, and which functionality can be safely hidden from casual users.

\subsection{Shared autonomy}

Another way to simplify the programming problem for end users is to take advantage of {\em shared autonomy}. The idea behind shared autonomy is to fall back on teleoperator assistance when autonomy fails, as it will invariably do for the near future. A teleoperator does not necessarily have to be the person who wrote the program. Instead, they could function similarly to staff in telephone call centers, where a team of trained professionals handle incidents as they arise and get the robots back on track to fulfill their programmed goals.

With a human expert on call to get a robot out of situations it was not programmed to handle, we can simplify the programming problem for end users so that they do not have to specify error recovery behaviors for every single corner case.
\end{changebar}

As a stepping-stone towards full autonomy, this approach also enables the testing of end user programmable systems without needing a fully reliable set of autonomous primitives in place to support them.

\begin{changebar}
\subsection{New 3D input devices}

Finally, video gaming technology is driving a new crop of input and output devices that make it easier to interact with and perceive the 3D world around us. These devices will make it easier for robots to perceive the world in three dimensions, and for people to instruct those robots using more natural gestures. Input devices such as the Razer Hydra, Leap Motion Controller, and Microsoft Kinect can sense motion in 3D, enabling computers to sense motion and objects in the world. 3D output devices such as the Oculus Rift project stereo images in 3D so that people can perceive what robots are seeing in three dimensions.

These new devices could enable a new generation of robot control interfaces that make interacting with robots more natural and intuitive.

\section{Conclusion}

Personal service robots have an enormous potential to do good for humanity by enabling people to live more independently. However, controlling these robots boils down to a small matter of programming. In this article I have explained why programming robots is inherently difficult, even for skilled programmers.  Drawing lessons from the field of end user programming, we have proposed several paths forward that may enable the creation of more accessible interfaces for personal robots in the future.
\end{changebar}

%Over time, as we refine existing autonomous primitives and build new domain-specific primitives, my hope is that we will expand the repertoire of robust end-user-programmable robot behavior. And perhaps that goal of having our own Rosie the Robot in our homes will soon become a reality.

% Acknowledgements
% Andreas Paepke
% Kaijen Hsiao
% Matei Ciocarlie
% Jon Binney
% Brandon Alexander
% Maya Cakmak
% anonymous reviewers
